---
title: "Analyze Your Data with R: A complete guide."
output:
  pdf_document:
    toc: true
---

\newpage

# 1 Introduction
## 1.1 What is _`tableone`_ package in _R_?
* `tableone` is an _R_ package that helps you create "Table 1" _(making summary statistics)_, as well as performing the basic needed statistical tests.
    
    
## 1.2 Features of _`tableone`_ Package:
*	Handles both continuous and categorical variables.
* Categorical variables are summarized as counts (how many) and/or percentages.
* Continuous variables can be summarized in two ways:
  + Normal way: Shows the mean (average) and standard deviation (how much values vary).
  + Non-normal way: Shows the median (middle value) and interquartile range (range of the middle 50% of values, excluding first and last quartiles).

\newpage
  
# 2 Getting into it
## 2.1 What Will You Need?
* To effectively utilize the _`tableone`_ package in _R_ for data analysis, the following tools and software are recommended:
  + **RStudio**: An integrated development environment (IDE) for _R_ that facilitates code writing, package usage, and data visualization.
  + **_`tableone`_ Package** itself.
  + **Spreadsheet Software**: Such as **Microsoft Excel** or any equivalent alternative, to export, review, and refine the generated tables.
  + **Word Processing Software**: Such as **Microsoft Word** or any similar application, to compile and finalize the analysis report, including formatted tables and interpretations.


## 2.2 How to Setup the Environment?
* Follow these steps to set up your environment:
  + **Install RStudio**: If you haven't already, [download and install RStudio](https://posit.co/download/rstudio-desktop/).
  + **Install Required Packages**: Once RStudio is installed and running, you'll need to install the necessary packages:
    - The _`tableone`_ package is essential for creating summary tables and finding p-values.
    - It is recommended to also install the _`tidyverse`_ library.

```{r eval=FALSE}
install.packages("tableone") # Installing the `tableone` package
install.packages("tidyverse") # Installing the `tidyverse` library

# Activating the packages
library(tableone)
library(tidyverse)
```

## 2.3 What about the Workflow?
1. **Import Your Data**: Usually from Excel file.
2. **Create Summary Tables**: Utilize the _`tableone`_ package to perform descriptive and inferential statistics.
3. **Export Tables**: Once your tables are ready, export each one to an Excel file.
4. **Integrate with Your Document**: Finally, copy the tables from Excel and paste them into your Word document.
\
\
_Detailed instructions for each step will be provided in the next sections._


\newpage


# 3 Usage
## 3.1 Importing the Raw Data from Excel File
* You can import the data file using RStudio interface:
  + Import Dataset _in the upper right corner_ => From Excel
* You can import it using r code:
```{r import}
my_data <- readxl::read_excel("my_data.xlsx")
# Storing the data in a variable called "my_data"



# NOTE: Some concepts you need to know
#
#   1. A variable: you can think of a variable as the pocket that will contain data
# during any process.
#
#   2. A function: you can think of a function as the machine that will take your data
# and operate on it, giving you an output based on the given data. The syntax of a
# function is: my_function(parameter, another_parameter, ..., another_parameters)
#
#   3. A parameter: It is a piece of information you pass to a function, it can be the
# data that the function will operate on, or it can be some instruction.
```
Here is a preview of the data we will work on during this tutorial:
```{r echo=FALSE}
my_data
```


## 3.2 Something to Put in Mind: _categorical variable conversion_
* Usually, categorical variables are coded numerically _(when working with R, it is better to leave it uncoded)_.
* Our package can't recognize this, So you have to include a list _(vector)_ with all the categorical variables of your data in a parameter _(instruction)_ while using the package.
_Will be clarified in the next section_

\newpage

## 3.3 Making Summary Statistics _(Descriptive Statistics)_
* Use the function **_'CreateTableOne()'_**

### 3.3.1 Simplest Use Case: treating the whole data as one group
* In this example, all columns of your data will be included.
```{r simple}
tableone::CreateTableOne(data = my_data)
# Here, we are using the function `CreateTableOne()` with the parameter "data" to 
# give it our data, which is stored in the variable "my_data"
```

### | 3.3.1.1 Choosing Only Some Variables to Be Included
* Here, we will add the parameter "vars":
```{r choose_vars}
## Vector of variables to summarize
included_vars <- c("Group", "Weight", "Height", "Hb") # Excluded the ID variable

## Create a TableOne
my_summary_table <- tableone::CreateTableOne(data = my_data, vars = included_vars)
# "vars" parameter determines the variables included in the summary.

print(my_summary_table) # The function "print" shows the table
```

\newpage

### | 3.3.1.2 Categorical Variable Conversion: Define categorical variables
* Here, we will add the parameter "factorVars" to determine which variables are categorical:
```{r define_cats}
## Vector of variables to summarize
included_vars <- c("Group", "Weight", "Height", "Hb") # Excluded the ID variable

## Vector of categorical variables that need transformation
categorical_vars <- c("Group")

## Create a TableOne
my_summary_table <- tableone::CreateTableOne(
  data = my_data,
  vars = included_vars,
  factorVars = categorical_vars
  )
# "factorVars" parameter determines the variables that should be treated as categorical.

print(my_summary_table) # The function "print" shows the table
```

\newpage

### 3.3.2 Showing all levels for categorical variables
* If you want to show all levels, you can use "showAllLevels" argument to the print() method in order to show all categories of the included categorical variables:
```{r show_all}
## Vector of variables to summarize
included_vars <- c("Group", "Weight", "Height", "Hb") # Excluded the ID variable

## Vector of categorical variables that need transformation
categorical_vars <- c("Group")

## Create a TableOne
my_summary_table <- tableone::CreateTableOne(
  data = my_data,
  vars = included_vars,
  factorVars = categorical_vars
  )

# Now we are using some new parameters for the print function
print(my_summary_table, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```

\newpage

### 3.3.3 Grouping the Data: dividing the data into groups according to given categorical variable
* Here, we will add the parameter "strata" and assign to it the categorical variable (grouping factor) by which the data will be grouped.
* You can also see that the package had perfomed the proper statistical test to find p-value of difference between groups of the given grouping factor.
```{r stratify}
## Vector of variables to summarize
included_vars <- c("Weight", "Height", "Hb") # Excluded the ID variable

## Vector of categorical variables that need transformation
categorical_vars <- c("Group")

## Create a TableOne
my_summary_table <- tableone::CreateTableOne(
  data = my_data,
  vars = included_vars,
  factorVars = categorical_vars,
  strata = "Group"
  )

# Now we are using some new parameters for the print function
print(my_summary_table, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```

\newpage

### 3.3.4 Summarizing Nonnormal Variables
* Do it with the nonnormal argument to the print() method:
```{r nonnormal}
## Vector of variables to summarize
included_vars <- c("Weight", "Height", "Hb") # Excluded the ID variable

## Vector of categorical variables that need transformation
categorical_vars <- c("Group")

# Lets assume that the height is not normally distributed
nonnormal_vars <- c("Height")

## Create a TableOne
my_summary_table <- tableone::CreateTableOne(
  data = my_data,
  vars = included_vars,
  factorVars = categorical_vars,
  strata = "Group"
  )

# Now we are using some new parameters for the print function
print(
  my_summary_table,
  nonnormal = nonnormal_vars,
  showAllLevels = TRUE,
  formatOptions = list(big.mark = ","))
```

\newpage

### 3.3.5 Detailed information including missingness
* If you need more detailed information including the number/proportion missing. Use the summary() method on the result object.
* It also shows the p-values in both cases of normality.
* The continuous variables are shown first, and the categorical variables are shown second.
* Unfortunately, our data doesn't contain categorical variables other than the grouping one, but if you try it with another data, it will work.

```{r detailed}
summary(my_summary_table)
```

\newpage


### 3.3.6 Categorical or Continuous Variables Only
* To get the categorical part only of the previous methods, you can use the following code before using "print":
```{r cat_only}
# We don't have categorical varaibles, but let's try the syntax!
my_summary_table <- my_summary_table$CatTable
```
* To get the continuous part only of the previous methods, you can use the following code before using "print":
```{r cont_only}
my_summary_table <- my_summary_table$ContTable
```

\newpage

## 3.4 Testing for Difference _(Finding the p-Value)_
* As you can see in the previous table, when there are two or more groups group comparison p-values are printed along with the table.
* The hypothesis test functions used by default are:
  + chisq.test() for categorical variables.
  + oneway.test() for continous variables (with equal variance assumption, i.e., regular ANOVA).
* You may be worried about the nonnormal variables and small cell counts in the stage variable.
* In such a situation, you can use the nonnormal argument like before as well as the exact (test) argument in the print() method.
  + Now kruskal.test() is used for the nonnormal continuous variables and fisher.test() is used for categorical variables specified in the exact argument.
```{r diff}
## Vector of variables to summarize
included_vars <- c("Weight", "Height", "Hb") # Excluded the ID variable

## Vector of categorical variables that need transformation
categorical_vars <- c("Group")

# Lets assume that the height is not normally distributed
nonnormal_vars <- c("Height")

## Create a TableOne
my_summary_table <- tableone::CreateTableOne(
  data = my_data,
  vars = included_vars,
  factorVars = categorical_vars,
  strata = "Group"
  )

# You test for the Standardize mean differences using the "smd" parameter
print(my_summary_table, nonnormal = nonnormal_vars, smd = TRUE)
```

\newpage

## 3.5 Exporting
* The following code exports the final table, which you made using the previous methods, into CSV file _(A type of data files, it is also supported by Microsoft Excel)_:
```{r eval=FALSE}
write.csv(tab, file = "myTable.csv")
# Save to a CSV file
```



\newpage
# 4 Summary
### In this tutorial, we explored the process of generating and customizing summary statistics tables using the _`tableone`_ package, with particular attention to the "`CreateTableOne()`" function. Below is a brief summary of the key aspects and modifications discussed:
* **Variable Selection and Stratification**:
  + You can specify variables to include using the "`vars`" argument.
  + You can stratify the table by categorical variables with the "`strata`" argument.
* **Showing All Levels for Categorical Variables**:
  + By adding 'showAllLevels = TRUE' to the print method.
* **Handling Non-Normal Distributions**:
  + The "`nonnormal`" argument allows the presentation of median and interquartile ranges (IQR) for non-normally distributed continuous variables.
* **Defining Factor Variables**:
    * The "`factorVars`" argument ensures proper handling of categorical data.
* **P-Value Calculation and Customization**:
    * p-values are automatically calculated using the proper tests when stratifying the data.

### This summary encapsulates the key modifications and parameters available in the "`CreateTableOne()`" function, offering a concise reference for efficiently generating and tailoring summary tables in your research work.

\newpage

# 5 Example: Place your parameters and run this code
```{r eval=FALSE}
# Ultimate Guide: analyze your data in simple steps
# ========================================================

# Importing the file
my_data <- readxl::read_excel("my file destination/my file.xlsx")

# Determining the included variables
included_vars <- c("included variable name 1", "included variable name n")
# Determining the nonnormal variables
nonnormal_vars <- c("nonnormal variable name 1", "nonnormal variable name n")
# Determining the categorical variables
categorical_vars <- c("categorical variable name 1", "categorical variable name n")
# Determining the grouping variable
grouping_variable <- "grouping variable name"

# Making the analysis
analysis_table <- CreateTableOne(
  data = my_data,
  vars = included_vars,
  factorVars = categorical_vars,
  strata = grouping_variable
)
final_table <- print(analysis_table,
      showAllLevels = TRUE,
      nonnormal = nonnormal_vars,
      formatOptions = list(big.mark = ","))

# Exporting the final_table to a CSV file
write.csv(final_table, file = "Data Analysis Results.csv")

# ========================================================
# Congrats! You've finished the analysis.
```